Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on

### 2011-06-03 paper.tex
Spellcheck

Note that while some may advocate also for Open Access to
publications, the Blue Obelisk goals (ODOSOS) focus more on the
availability of the underlying scientific data, standards (to exchange
data), and code (to reproduce results). All three of these goals stem
from the fundamental tenants of the scientific method for data sharing
and reproducibilty.

@@

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse is an award-winning
molecular workbench for life sciences
that wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
preparing and analysing computational chemistry calculations. Both
projects are designed to be extended or scripted by users through
the provision of a plugin architective and scripting support (using
Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
of Avogadro). An interesting aspect of both Avogadro and Bioclipse is
that they share some developers with the underlying toolkits and this
has driven the development of new features in the CDK and Open Babel.

@@

Another area where the Blue Obelisk has had a signficant impact in the
past five years is in supporting
quantum chemistry calculations and in interpreting their results.
Electronic structure calculations have a long tradition in the
chemistry community and a variety of programs exist, mostly
proprietary software with the notable exception of
NWChem~\cite{NWChem}.
However, since each program uses different input formats, and the
the output formats vary widely (sometimes even varying between different versions
of the same software), preparing calculations and automatically extracting
the results is problematic.

@@

In 2006, the Blue Obelisk project cclib~\cite{cclib} was established
with the goal of parsing the output from computational chemistry
programs and presenting it in a standard way so that further analyses
could be carried out independently of the quantum package used.
cclib is a Python library, and the current version (version 1.0.1)
supports 8 different computational chemistry codes and extracts over
30 different calculated attributes. Two related BO projects build upon 
cclib. GaussSum~\cite{WebGaussSum},
is a GUI that can monitors the progress of SCF and geometry convergences, 
and can plot predicted UV/Vis absorption and infrared spectra from 
appropriate logfiles containing energies and oscillator strengths for easy 
comparison to experimental data. QMForge~\cite{WebQMForge} provides 
a GUI for various electronic structure analyses such as Frenking's charge 
decomposition analysis~\cite{Frenking} and Mulliken or C-squared analyses
on user-defined molecular fragments. QMForge also provides a rudamentary
Cartesian coordinate editor allowing molecular structures to be saved via Open Babel.

@@

Industry participants also pointed out that 
there is a considerable amount of
contribution-in-kind from industry, both from
enhancements to software and also the development of completely new
software and toolkits. Companies are now finding it easier to create
mechanisms for releasing Open Source software
without violating confidentiality or incurring liability.
A phrase from the meeting
summed it up: ``The ice is beginning to melt'', signifiying that we can
expect a rapid increase in industry's interest
in Open Source.

@@

Besides the traditional and proven relational database approach with
added chemical features (`cartridges'), there is growing
interest in tools and approaches based on the web philosophy and
practice. Several groups are experimenting with 
the Resource Description Framework (RDF) language
on the assumption that generic high-performance solutions will appear.
RDF allows everything to be described by
URIs (data, molecules, dictionaries, relations). The Chempound
system~\cite{Chempound},
as deployed in
Quixote and elsewhere, is an RDF-based approach to chemical structures
and compounds and their properties. For small
to medium-sized collections (such as an individual's calculations or
literature retrieval), there are many RDF tools
(e.g. SIMILE, Apache Jena) which can operate in machine memory and provide
the flexibility that RDF offers. For larger
systems, it is unclear whtether complete RDF solutions (e.g. Virtuoso)
will be satisfactory or whether a hybrid system
based on name-value pairs (e.g. CouchDB, MongoDB) will be sufficient.

>>>>

Note that while some may advocate also for Open Access to
publications, the Blue Obelisk goals (ODOSOS) focus more on the
availability of the underlying scientific data, standards (to exchange
data), and code (to reproduce results). All three of these goals stem
from the fundamental tenants of the scientific method for data sharing
and reproducibility.

@@

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse is an award-winning
molecular workbench for life sciences
that wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
preparing and analysing computational chemistry calculations. Both
projects are designed to be extended or scripted by users through
the provision of a plugin architecture and scripting support (using
Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
of Avogadro). An interesting aspect of both Avogadro and Bioclipse is
that they share some developers with the underlying toolkits and this
has driven the development of new features in the CDK and Open Babel.

@@

Another area where the Blue Obelisk has had a significant impact in the
past five years is in supporting
quantum chemistry calculations and in interpreting their results.
Electronic structure calculations have a long tradition in the
chemistry community and a variety of programs exist, mostly
proprietary software with the notable exception of
NWChem~\cite{NWChem}.
However, since each program uses different input formats, and the
the output formats vary widely (sometimes even varying between different versions
of the same software), preparing calculations and automatically extracting
the results is problematic.

@@

In 2006, the Blue Obelisk project cclib~\cite{cclib} was established
with the goal of parsing the output from computational chemistry
programs and presenting it in a standard way so that further analyses
could be carried out independently of the quantum package used.
cclib is a Python library, and the current version (version 1.0.1)
supports 8 different computational chemistry codes and extracts over
30 different calculated attributes. Two related BO projects build upon 
cclib. GaussSum~\cite{WebGaussSum},
is a GUI that can monitors the progress of SCF and geometry convergences, 
and can plot predicted UV/Vis absorption and infrared spectra from 
appropriate logfiles containing energies and oscillator strengths for easy 
comparison to experimental data. QMForge~\cite{WebQMForge} provides 
a GUI for various electronic structure analyses such as Frenking's charge 
decomposition analysis~\cite{Frenking} and Mulliken or C-squared analyses
on user-defined molecular fragments. QMForge also provides a rudimentary
Cartesian coordinate editor allowing molecular structures to be saved via Open Babel.

@@

Industry participants also pointed out that 
there is a considerable amount of
contribution-in-kind from industry, both from
enhancements to software and also the development of completely new
software and toolkits. Companies are now finding it easier to create
mechanisms for releasing Open Source software
without violating confidentiality or incurring liability.
A phrase from the meeting
summed it up: ``The ice is beginning to melt'', signifying that we can
expect a rapid increase in industry's interest
in Open Source.

@@

Besides the traditional and proven relational database approach with
added chemical features (`cartridges'), there is growing
interest in tools and approaches based on the web philosophy and
practice. Several groups are experimenting with 
the Resource Description Framework (RDF) language
on the assumption that generic high-performance solutions will appear.
RDF allows everything to be described by
URIs (data, molecules, dictionaries, relations). The Chempound
system~\cite{Chempound},
as deployed in
Quixote and elsewhere, is an RDF-based approach to chemical structures
and compounds and their properties. For small
to medium-sized collections (such as an individual's calculations or
literature retrieval), there are many RDF tools
(e.g. SIMILE, Apache Jena) which can operate in machine memory and provide
the flexibility that RDF offers. For larger
systems, it is unclear whether complete RDF solutions (e.g. Virtuoso)
will be satisfactory or whether a hybrid system
based on name-value pairs (e.g. CouchDB, MongoDB) will be sufficient.
